---
title: "Reproducible Research - Course Project 1"
author: "Danielle Boucher"
date: "November 27, 2016"
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Activity Monitoring Data
### Loading and preprocessing the data

In this project, we will be working with the **activity monitoring data set** available [here.] [1] This data set has been unzipped and saved to our active working directory. Next, we will read the table into R, and take a look at it:  
  
```{r read}
activitydata <- read.csv("activity.csv")
head(activitydata)
str(activitydata)
```

Great! Looks like this data set is pretty straightforward - 3 dimensions and 17568 observations. Of course, it makes more sense to convert **date** to a date:  
```{r convert}
activitydata$date <- as.Date(activitydata$date)
str(activitydata)
```

We're also going to want some summarized data sest, to view the total number of steps taken by date, and the average number of steps by interval:  
```{r summed}
activitysum <- aggregate(activitydata$steps ~ activitydata$date, FUN=sum)
head(activitysum)
intervalmean <- aggregate(activitydata$steps ~ activitydata$interval, FUN=mean)
head(intervalmean)
```

### What is mean total number of steps taken per day?
Let's take a look at some of the key attributes of the summary data.  
We'll make a *histogram* and check the *mean* and *median*:
```{r total}
hist(activitysum$`activitydata$steps`, col="RED", xlab="Steps", main="Freq. of Total Steps")
mean(activitysum$`activitydata$steps`, na.rm=TRUE)
median(activitysum$`activitydata$steps`, na.rm=TRUE)
```


### What is the average daily activity pattern?
Great! Next we'll try to better understand average activity by day, using the **intervalmean** summary we created.  

```{r intervals}
plot(intervalmean$`activitydata$interval`, intervalmean$`activitydata$steps`, type="l", xlab="Interval", ylab="Average Steps", main="Average Steps by Time of Day Interval")
intervalmean[which.max(intervalmean$`activitydata$steps`),]
```

Okay, so we can see how the number of steps taken throughout the day looks, on average. We can also see that interval 835 has the maximum average steps, at 206.  

### Imputing missing values
Let's move on. The data set has a number of missing values that may be impacting our results. It would be valuable to try and remove this bias. Here's some code from replacing the missing values with the average for that interval:  

```{r missingno}
## How many missing values for steps are there? 
summary(is.na(activitydata$steps))
activitydataclean <- activitydata
activitydataclean$steps[is.na(activitydataclean$steps)] <- with(activitydataclean, ave(steps, interval, FUN = function(x) mean(x, na.rm = TRUE)))[is.na(activitydataclean$steps)]
str(activitydataclean)
## Confirm that we replaced all
summary(is.na(activitydataclean$steps))
```

Okay! Comparing that to our first *str()* call from the beginning of this document, we can see that we've successfully replaced the NAs with average values for the interval in question! Let's see now how our summary data compares:

```{r newsummary}
activitysum2 <- aggregate(activitydataclean$steps ~ activitydataclean$date, FUN=sum)
hist(activitysum2$`activitydataclean$steps`, col="BLUE", xlab="Steps", main="Freq. of Total Steps")
mean(activitysum2$`activitydataclean$steps`, na.rm=TRUE)
median(activitysum2$`activitydataclean$steps`, na.rm=TRUE)
```

The shape of our histogram has stayed the same, but the frequency of each bucket has increased. The mean and median have not changed much at all. This tells me that we've probably chosen an acceptable method of cleaning the missing values.  

### Are there differences in activity patterns between weekdays and weekends?
Moving on... Let's see what correlations exist between steps taken, and day of the week. 

```{r weekdays}
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
activitydataclean$weekday <- factor((weekdays(activitydataclean$date) %in% weekdays1), levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))
summary(activitydataclean$weekday)
weekenddata <- subset(activitydataclean, activitydataclean$weekday == "weekday")
weekenddata <- aggregate(weekenddata$steps ~ weekenddata$interval, FUN=mean)
weekdaydata <- subset(activitydataclean, activitydataclean$weekday == "weekend")
weekdaydata <- aggregate(weekdaydata$steps ~ weekdaydata$interval, FUN=mean)
par(mfcol=c(2,1))
plot(weekenddata$`weekenddata$steps` ~ weekenddata$`weekenddata$interval`, type="l", ylab="Steps", xlab="Interval", main="Weekends")
plot(weekdaydata$`weekdaydata$steps` ~ weekdaydata$`weekdaydata$interval`, type="l", ylab="Steps", xlab="Interval", main="Weekdays")
```

Okay... interesting results! That's all we'll look at in the activity data set, for now!

[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip
